|

### 2023-04-03 appendix.Rmd
rearranged main paper and appendix into separate files again

---
title: |
  A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol
type: ARTICLE TEMPLATE
author:
  - name: Weihao Li
    affil: a
    email: weihao.li@monash.edu
  - name: Dianne Cook
    affil: a
    email: dicook@monash.edu
  - name: Emi Tanaka
    affil: a
    email: emi.tanaka@monash.edu
  - name: Susan VanderPlas
    affil: b
    email: susan.vanderplas@unl.edu
affiliation:
  - num: a
    address: |
      Department of Econometrics and Business Statistics, Monash University, Clayton, VIC, Australia
  - num: b
    address: |
      Department of Statistics, University of Nebraska, Lincoln, Nebraska, USA
bibliography: paper.bib
abstract: |
  Abstract to fill.
keywords: |
  statistical graphics; data visualization; visual inference; hypothesis testing; reression analysis; cognitive perception; simulation; practical significance; effect size
header-includes: |
  \usepackage{lscape}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
output: rticles::tf_article
editor_options: 
  chunk_output_type: console
---

@@

# Introduction

> *"Since all models are wrong the scientist must be alert to what is importantly wrong."* [@box1976science]

Diagnostics are the key to determining whether there is anything importantly wrong with a model. In linear regression analysis, residuals from the model fit are commonly used. Residuals summarise what is not captured by the model, and thus provide the capacity to identify what might be wrong.

We can assess residuals in multiple ways. Residuals may be plotted as a histogram or quantile-quantile plot to examine the distribution. Using the classical normal linear regression model as an example, if the distribution is symmetric and unimodal, we consider it to be well behaved. However, if the distribution is skewed, bimodal, multimodal, or contains outliers, there is a cause for concern. One could also inspect the distribution by conducting a goodness of fit test, such as the Shapiro-Wilk Normality test [@shapiro1965analysis].

More typically, residuals will be plotted, as a scatter plot against the predicted values and each of the explanatory variables to scrutinize their relationships. If there are any visually discoverable patterns, the model is potentially misspecified. In general, one looks for noticeable departures from the model like non-linear dependency or heteroskedasticity. However, correctly judging a residual plot where no pattern exists can be a painstakingly difficult task for humans. It is especially common, particularly among students, to misinterpret patterns that are random noise and random deviation from a model [@loy2021bringing]. It is also possible to conduct hypothesis tests for non-linear dependence [@ramsey_tests_1969], and use a Breusch-Pagan test [@breusch_simple_1979] for heteroskedasticity.

Abundance of literature describe appropriate diagnostic methods for linear regression, e.g. @draper1998applied, @montgomery1982introduction, @belsley_regression_1980, @cook_applied_1999 and @cook1982residuals. All these writings consider plotting residuals to be a standard technique that should be examined routinely in all regression modelling problems. In addition, @draper1998applied and @belsley_regression_1980 believe that residual plots are usually revealing when the assumptions are violated. @cook_applied_1999 thinks formal tests and graphical procedures are complementary and both have a place in residual analysis, but they focus on graphical methods rather than on formal testing, as they are easier to use. @montgomery1982introduction even suggests that residual plots are more informative in most practical situations than the corresponding formal tests, and statistical tests on regression model residuals are not widely used based on their experience.

A common guidance by experts is that optimal method for diagnosing model fits is by plotting the data. The persistence of this advice to check the plots is curious, and investigating why this might be common advice is the subject of this paper. The paper is structured as follows. The next section describes the background on the types of departures that one expects to detect, and outlines a formal statistical process for reading residual plots, called visual inference. Section \ref{experimental-design} details the experimental design to compare the decisions made by formal hypothesis testing, and how humans would read diagnostic plots. The results are reported in Section \ref{results}. We conclude with a discussion of future work, in particular, how the responsibility for residual plot reading might be relieved.

# Background

## Departures from good residual plots

```{r residual-plot-common-departures, fig.width = 10, fig.height = 2.5, fig.cap = "Example residual vs fitted value plots: (A) classically good looking residuals, (B) non-linear pattern indicates that the model has not captured a non-linear association, (C) heteroskedasticity indicating that variance around the fitted model is not uniform, and (D) non-normality where the residual distribution is not symmetric around 0. The latter pattern might best be assessed using a univariate plot of the residuals, but patterns B and C need to be assessed using a residual vs fitted value plot."}

set.seed(10086)

# Base model
mod <- poly_model(include_z = FALSE, sigma = 0.25)

# Base data
dat <- mod$gen(300)

bind_rows(
  dat %>%
    mutate(type = "(A) Good residuals"),
  mod$set_prm("include_z", TRUE)$
    gen(300, computed = select(dat, x, e)) %>%
    mutate(type = "(B) Non-linearity"),
  heter_model(b = 64)$
    gen(300, computed = select(dat, x, e)) %>%
    mutate(type = "(C) Heteroskedasticity"),
  heter_model(b = 0, e = rand_lognormal(sigma = 0.5))$
    gen(300, computed = select(dat, x)) %>%
    mutate(type = "(D) Non-normality")
) %>%
  mod$plot(remove_axis = TRUE, remove_grid_line = TRUE, theme = theme_light()) +
  facet_wrap(~type, ncol = 4, scales = "free") +
  xlab("Fitted values") +
  ylab("Residuals")
```

Graphical summaries in which residuals are plotted against fitted values or other functions of the predictors that are approximately orthogonal to the residuals are referred to as standard residual plots in @cook1982residuals. Figure \ref{fig:residual-plot-common-departures}A shows an ideal residual plot where the residuals are evenly distributed at both sides of the horizontal zero line, with no noticeable patterns.

There are various types of departures from an ideal residual plot. Non-linearity, heteroskedasticity and non-normality are perhaps the three mostly checked departures.

Non-linearity is a type of model misspecification caused by failing to include higher order terms of the predictors in the regression equation. Any non-linear functional form of residuals on fitted values in the residual plot could be indicative of non-linearity. An example residual plot containing visual pattern of non-linearity is shown in Figure \ref{fig:residual-plot-common-departures}B. One can clearly observe the "S-shape" from the residual plot as the cubic term is not captured by the misspecified model.

Heteroskedasticity refers to the presence of nonconstant error variance in a regression model. It is mostly due to the strict but false assumptions on the variance-covariance matrix of the error term. The usual pattern of heteroskedasticity on a residual plot is the inconsistent spread of the residuals across the horizontal axis. Visually, it sometimes results in the so-called "butterfly" shape as shown in Figure \ref{fig:residual-plot-common-departures}C, or the "left-triangle" and "right-triangle" shape where the smallest variance occurs at one side of the horizontal axis.

Compared to non-linearity and heteroskedasticity, non-normality is usually harder to detect from a residual plot since a scatter plot do not readily reveal the marginal distribution. A favourable graphical summary for this task is the quantile-quantile plot. As we mainly discuss residual plots, non-normality will not be the focus of this paper. For a consistent comparison, the residual plot of this departure is still presented in Figure \ref{fig:residual-plot-common-departures}D. When the number residuals below and above the horizontal axis are uneven across the local regions along the $x$-axis, we expect that the normality assumption is violated. For example, given a skewed error distribution, there will be fewer data points and more outliers on one side of the horizontal axis as shown in Figure \ref{fig:residual-plot-common-departures}D.

## Conventionally testing for departures

Other than checking diagnostic plots, analysts may perform formal hypothesis testing to detect model defects. A variety of tests can be applied; each testing for a specific violation of the null hypothesis. For example, the presence of heteroskedasticity can usually be tested by applying the White test [@white_heteroskedasticity-consistent_1980] or the Breusch-Pagan test [@breusch_simple_1979], which are both derived from the Lagrange multiplier test [@silvey1959lagrangian] principle that relies on the asymptotic properties of the null distribution. To test specific forms of non-linearity, one may apply the F-test as a model structural test to examine the significance of specific polynomial and non-linear forms of the predictors, or the significance of proxy variables as in the Ramsey Regression Equation Specification Error Test (RESET) [@ramsey_tests_1969]. The Shapiro-Wilk test [@shapiro1965analysis] is the most widely used test of non-normality included by many of the statistical software programs. The Jarque--Bera test [@jarque1980efficient] is also used to directly check whether the sample skewness and kurtosis match a normal distribution.

We apply the RESET test, Breusch-Pagan test and Shapiro-Wilk test to the residual plots shown in Figure \ref{fig:residual-plot-common-departures}; the results of this is shown in Table \ref{tab:example-residual-plot-table}. The Breusch-Pagan test and the Shapiro-Wilk test both reject $H_0$ for residual plots that display non-linearity and heteroscedasticity, even though these are not the original intention of the test. As discussed in @cook1982residuals, most residual-based tests for a particular type of departure from model assumptions are also sensitive to other types of departures. It is likely $H_0$ is correctly rejected but for the wrong reason, a phenomenon known as the "Type III error". Additionally, outliers will often incorrectly trigger the rejection of $H_0$ despite when majority of the residuals are well-behaved [@cook_applied_1999]. Furthermore, with a sufficiently large sample size, residual-based tests may reject $H_0$ due to a slight departure that is of little practical significance. These can be largely avoided in diagnostic plots as experienced analysts can evaluate the acceptability of assumptions flexibly, even in the presence of outliers and slight departures.

```{r}
set.seed(10086)

# Base model
mod <- poly_model(include_z = FALSE, sigma = 0.25)

# Base data
dat0 <- mod$gen(300)

# Replicate data in Figure 1
dat1 <- mod$set_prm("include_z", TRUE)$gen(300, computed = select(dat0, x, e))
dat2 <- heter_model(b = 64)$gen(300, computed = select(dat0, x, e))
dat3 <- heter_model(b = 0, e = rand_lognormal(sigma = 0.5))$
               gen(300, computed = select(dat0, x))

dat_list <- list(dat0, dat1, dat2, dat3)

table_dat <- data.frame(plot = c("A", "B", "C", "D"), 
                        model = c("None", 
                                  "Non-linearity", 
                                  "Heteroskedasticity", 
                                  "Non-normality"),
                        r = map_dbl(dat_list, 
                                    ~POLY_MODEL$test(.x, 
                                                     test = "RESET", 
                                                     power = 2:4)$p_value),
                        b = map_dbl(dat_list, 
                                    ~HETER_MODEL$test(.x)$p_value),

                        # SW test has not been included by visage
                        s = map_dbl(dat_list, 
                                    ~shapiro.test(.x$.resid)$p.value)) %>%

  # 3 decimal points
  mutate(across(r:s, ~ format(round(.x, digits = 3), nsmall = 3))) %>%

  # Italic if reject
  mutate(across(r:s, ~ kableExtra::cell_spec(.x, italic = as.numeric(.x) <= 0.05)))

  table_dat %>%
  knitr::kable(col.names = c("Plot", 
                             "Departures", 
                             "RESET", 
                             "Breusch-Pagan", 
                             "Shapiro–Wilk"),
               align = "llrrr",
               escape = FALSE,
               format = "latex",
               booktabs = TRUE, 
               caption  = 'Statistical significance testing for departures from good residuals for plots in Figure \\ref{fig:residual-plot-common-departures}. Shown are the $p$-values calculated for the RESET, the Breusch-Pagan and the Shapiro–Wilk tests. The good residual plot (A) is judged a good residual plot, as expected, by all tests. The non-linearity (B) is detected by all tests, as might be expected given the extreme structure.', 
               label = "example-residual-plot-table" ) 
```

## Visual test procedure based on lineups

One may argue that reading diagnostic plots is to some extent subjective and indecisive compared to those rigorous statistical procedures as it relies on graphical perception - human ability to interpret and decode the information embedded in graph [@cleveland_graphical_1984]. Further, the degree of the presence of the visual features typically can not be measured quantitatively and objectively, which may lead to over or under-interpretations of the data. For instance, people 
over-interpret the separation between gene groups in a two-dimensional projection from a linear discriminant analysis when in fact there are no differences in the expression levels between the gene groups and separation is not an uncommon occurrence [@roy_chowdhury_using_2015].

Visual inference was first introduced in a 1999 Joint Statistical Meetings (JSM) talk with the title "Inference for Data Visualization" by @buja_inference_1999 as an idea to address the issue of valid inference for visual discoveries of data plots. Later, @buja_statistical_2009 proposed the lineup protocol as a visual test inspired by the "police lineup" or "identity parade" which is the act of asking the eyewitness to identify criminal suspect from a group of irrelevant people. The protocol consists of $m$ randomly placed plots, where one plot is the data plot, and the remaining $m - 1$ null plots have the identical graphical procedure except the data has been replaced with data consistent with $H_0$, that the model is correctly specified. Then, an observer who have not seen the data plot will be asked to point out the most different plot from the lineup. Under $H_0$, it is expected that the data plot would have no distinguishable difference from the null plots, and the probability that the observer correctly picks the data plot is $1/m$. If one rejects $H_0$ as the observer correctly picks the data plot, then the Type I error of this test is $1/m$.

```{r first-example-lineup, fig.height = 7, fig.width = 7, fig.cap = "Visual testing is conducted using a lineup, as in the example here. The residual plot computed from the observed data (plot $2^2 + 2$, exhibiting non-linearity) is embedded among 19 null plots, where the residuals are simulated from a standard error model. Computing the $p$-value requires that the lineup be examined by a number of human judges, each asked to select the most different plot. A small $p$-value would result from a substantial number selecting plot $2^2 + 2$."}

vi_lineup$poly_300$data %>%
  VI_MODEL$plot_lineup(theme = theme_light(), remove_axis = TRUE, remove_grid_line = TRUE)
```

Figure \ref{fig:first-example-lineup} is an example of a lineup protocol. If the data plot at position $2^2 + 2$ is identifiable, then it is evidence for the rejection of $H_0$. In fact, the actual residual plot is obtained from a misspecified regression model with missing non-linear terms.

Data used in the $m - 1$ null plots needs to be simulated. In regression diagnostics, sampling data consistent with $H_0$ is equivalent to sampling data from the assumed model. As @buja_statistical_2009 suggested, $H_0$ is usually a composite hypothesis controlled by nuisance parameters. Since regression models can have various forms, there is no general solution to this problem, but it sometimes can be reduced to a so called "reference distribution" by applying one of the three methods: (i) sampling from a conditional distribution given a minimal sufficient statistic under $H_0$, (ii) parametric bootstrap sampling with nuisance parameters estimated under $H_0$, and (iii) Bayesian posterior predictive sampling. The conditional distribution given a minimal sufficient statistic is the best justified reference distribution among the three [@buja_statistical_2009]. Essentially, null residuals can be simulated by regressing $N$ i.i.d standard normal random draws on the predictors, then rescaling it by the ratio of residual sum of square in two regressions.

The effectiveness of lineup protocol for regression analysis is validated by @majumder_validation_2013 under relatively simple settings with up to two predictors. Their results suggest that visual tests are capable of testing the significance of a single predictor with a similar power as a t-test, though they express that in general it is unnecessary to use visual inference if there exists a conventional test, and they do not expect the visual test to perform equally well as the conventional test. In their third experiment, where there is not a conventional test, visual test outperforms the conventional test for a large margin. This is encouraging, as it promotes the use of visual inference in situations where there are no existing statistical testing procedures. Visual inference have also been integrated into diagnostic of hierarchical linear models by @loy2013diagnostic, @loy2014hlmdiag and @loy2015you. They use lineup protocols to judge the assumption of linearity, normality and constant error variance for both the level-1 and level-2 residuals.

# Calculation of statistical significance and test power

## What is being tested?

In diagnosing a model fit from residuals, we are generally interested in $H_0:$ *The regression model is correctly specified* against the broad alternative $H_a:$ *The regression model is misspecified*.

However, it is practically impossible to test this specific $H_0$ with conventional tests, which are constructed to measure specific departures. For example, the RESET test is formulated as $H_0:\gamma_1 = \gamma_2 = \gamma_3 = 0$ against $H_a: \gamma_1 \neq 0 \text{ or } \gamma_2 \neq 0 \text{ or } \gamma_3 \neq 0$, from $y = \tau_0 + \sum_{i=1}^{p}\tau_px_p +\gamma_1\hat{y}^2 + \gamma_2\hat{y}^3 + \gamma_3\hat{y}^4 + u, ~~u \sim N(0, \sigma_u^2)$. 
Similarly, the Breusch-Pagan test is designed to specifically test $H_0:$ *error variances are all equal* ($\zeta_i=0 \text{ for } i=1,..,p$) versus the alternative $H_a:$ *that the error variances are a multiplicative function of one or more variables* ($\text{at least one } \zeta_i\neq 0$) from $e^2 = \zeta_0 + \sum_{i=1}^{p}\zeta_i x_i + u, ~ u\sim N(0,\sigma_u^2)$.

One of the potential benefits of the visual test, based on the lineup protocol, is that it should be able to detect a range of departures from good residuals.

## Statistical significance

In hypothesis testing, a $p$-value is defined as the probability of observing test results as least as extreme as the observed result given $H_0$ is true. Conventional hypothesis tests usually have an existing method to derive or compute $p$-value based on the null distribution. What we need to discuss in the following is the method to estimate $p$-value for a visual test.

Within the context of visual inference, by involving $k$ independent observers, the visual $p$-value can be interpreted as the probability of having as many or more subjects detect the data plot than the observed result.

Let $X_j = \{0,1\}$ be a Bernoulli random variable denoting whether subject $j$ correctly detecting the data plot, and $X = \sum_{j=1}^{K}X_j$ be the number of observers correctly picking the data plot. Then, by imposing a relatively strong assumption on the visual test that all $K$ evaluations are fully independent, under $H_0$, $X \sim \mathrm{Binom}_{K,1/m}$. Therefore, the $p$-value of a lineup of size $m$ evaluated by $K$ observer is given as $P(X \geq x) = 1 - F(x) + f(x)$, where $F(.)$ is the binomial cumulative distribution function, $f(.)$ is the binomial probability mass function and $x$ is the realization of number of observers correctly picking the data plot [@majumder_validation_2013].

As pointed out by @vanderplas2021statistical, this basic binomial model does not take into account the possible dependencies in the visual test due to repeated evaluations of the same lineup. And it is inapplicable to visual test where subjects are asked to select one or more "most different" plots from the lineup. @vanderplas2021statistical summarises three common scenarios in visual inference: (1) $K$ different lineups are shown to $K$ subjects, (2) $K$ lineups with different null plots but the same data plot are shown to $K$ subjects, and (3) the same lineup is shown to $K$ subjects. Out of these three scenarios, Scenario 3 is the most common in previous studies as it puts the least constraints on the experiment design. For Scenario 3, @vanderplas2021statistical models the probability of a plot $i$ being selected from a lineup as $\theta_i$, where $\theta_i \sim Dirichlet(\alpha)$ for $i=1,...,m$ and $\alpha > 0$. The number of times plot $i$ being selected in $K$ evaluations is denoted as $c_i$. In case subject $j$ makes multiple selections, $1/s_j$ will be added to $c_i$ instead of one, where $s_j$ is the number of plots subject $j$ selected for $j=1,...K$. This ensures $\sum_{i}c_i=K$. Since we are only interested in the selections of the data plot $i$, the marginal model can be simplified to a beta-binomial model and thus the visual $p$-value is given as

```{=tex}
\begin{equation} \label{eq:pvalue-beta-binomial}
P(C \geq c_i) = \sum_{x=c_i}^{K}{K \choose x}\frac{B(x + \alpha, K - x + (m - 1)\alpha)}{B(\alpha, (m-1)\alpha)},\quad \text{for} \quad c_i \in \mathbb{Z}_0^+
\end{equation}
```

\noindent where $B(.)$ is the beta function defined as

```{=tex}
\begin{equation} \label{eq:betafunction}
B(a, b) = \int_{0}^{1}t^{\alpha - 1}(1-t)^{b-1}dt,\quad \text{where}\quad a,b>0.
\end{equation}
```

Note that Equation \ref{eq:pvalue-beta-binomial} given in @vanderplas2021statistical only works with non-negative integer $c_i$. We extend the equation to non-negative real number $c_i$ by applying a linear approximation

```{=tex}
\begin{equation} \label{eq:pvalue-beta-binomial-approx}
P(C \geq c_i) = P(C \geq \lceil c_i \rceil) + (\lceil c_i \rceil - c_i) P(C = \lfloor c_i \rfloor), \quad \text{for}\quad c_i \in \mathbb{R}_0^+,
\end{equation}
```

where $P(C \geq \lceil c_i \rceil)$ is calculated using Equation \ref{eq:pvalue-beta-binomial} and $P(C = \lfloor c_i \rfloor)$ is calculated by

\begin{equation} \label{eq:pmf-beta-binomial}
P(C = c_i) = {K \choose c_i}\frac{B(c_i + \alpha, K - c_i + (m - 1)\alpha)}{B(\alpha, (m-1)\alpha)},\quad \text{for} \quad c_i \in \mathbb{Z}_0^+.
\end{equation}

Besides, the parameter $\alpha$ used in Equation \ref{eq:pvalue-beta-binomial} and \ref{eq:pmf-beta-binomial} is usually unknown and hence needs to be estimated from the survey data. For low values of $\alpha$, only a few plots are attractive to the observers and tend to be selected. For higher values of $\alpha$, the distribution of the probability of each plot being selected is more even. @vanderplas2021statistical defines that a plot is $c$-interesting if $c$ or more participants select the plot as the most different. Given the definition, The expected number of plots selected at least $c$ times, $E[Z_c]$, is calculated as

```{=tex}
\begin{equation} \label{eq:c-interesting-expectation}
E[Z_c(\alpha)] = \frac{m}{B(\alpha, (m-1)\alpha)}\sum_{\lceil c \rceil}^{K}{K \choose x} B(x + \alpha, K - x + (m-1)\alpha).\end{equation}
```

With Equation \ref{eq:c-interesting-expectation}, $\alpha$ can be estimated using maximum likelihood estimation. But for precise estimate of $\alpha$, additional human responses to Rorschach lineups, which is a type of lineup that consists of plots constructed from the same null data generating mechanism, are required.

## Power of the tests

The power of a model misspecification test is the probability that $H_0$ is rejected given the regression model is misspecified in a specific way. It is an important indicator
when one is concerned about whether model assumptions have been violated. Although in practice, one might be more interested in knowing how much the residuals deviate from the model assumptions, and whether this deviation is of practical significance.

The power of a conventional hypothesis test is affected by both the true parameter $\boldsymbol{\theta}$ and the sample size $n$. These two can be quantified in terms of effect size $E$ to measure the strength of the residual departures from the model assumptions. Details about the effect size is provided in Section \ref{effect-size} after the introduction of the simulation model used in our human subject experiment. The theoretical power of a test is sometimes not a trival solution, but it can be estimated if the data generating process is known. We use a predefined model to generate a large set of simulated data under different effect sizes, and record if the conventional test rejects $H_0$. The probability of the conventional test rejects $H_0$ is then fitted by a logistic regression formulated as

```{=tex}
\begin{equation} \label{eq:logistic-regression-1-1}
Pr(\text{reject}~H_0|H_1,E) = \Lambda\left(log\left(\frac{0.05}{0.95}\right) + \beta_1 E\right),
\end{equation}
```

\noindent where $\Lambda(.)$ is the standard logistic function given as $\Lambda(z) = exp(z)/(1+exp(z))$. The effect size $E$ is the only predictor and the intercept is fixed to $log(0.05/0.95)$ so that $\hat{Pr}(\text{reject}~H_0|H_1,E = 0) = 0.05$, which is the desired significance level.

The power of a visual test on the other hand, may additionally depend on the ability of the particular subject, as the skill of the individual may affect the number of observers who identify the data plot from the lineup [@majumder_validation_2013]. To address this issue, @majumder_validation_2013 models the probability of a subject $j$ correctly picking the data plot from a lineup $l$ using a mixed-effect logistic regression, with subjects treated as random effect. Then, the estimated power of a visual test evaluated by a single subject is the predicted value obtained from the mix-effect model. However, this mix-effect model does not work with scenario where subjects are asked to select one or more most different plots. In this scenario, having the probability of a subject $j$ correctly picking the data plot from a lineup $l$ is insufficient to determine the power of a visual test because it does not provide information about the number of selections made by the subject for the calculation of the $p$-value (See Equation \ref{eq:pvalue-beta-binomial-approx}). Therefore, we directly estimate the probability of a lineup being rejected by assuming that individual skill has negligible effect on the variation of the power. This assumption is not necessary true, but it helps simplifying the model structure, thereby obviate a costly large-scale experiment to estimate complex covariance matrices. The same model given in Equation \ref{eq:logistic-regression-1-1} is applied to model the power of a visual test.

To study various factors contributing to the power of both tests, the same logistic regression model is fit on different subsets of the collated data grouped by levels of factors. These include the distribution of the fitted values, type of the simulation model and the shape of the residual departures.

# Experimental design

An experiment is conducted in three data collection periods to investigate the difference between conventional hypothesis testing and visual inference in the application of linear regression diagnostics. Two types of departures, non-linearity and heteroskedasticity, are collected during data collection periods I and II. The data collection period III was designed primarily to measure human responses to null lineups so that the parameter $\alpha$ in Equation \ref{eq:pvalue-beta-binomial} can be estimated. Additional lineups for both non-linearity and heteroskedasticity, using uniform fitted value distribution, were included so that the participants were evaluating some lineups with signal also. It would be too frustrating for participants to only be assigned lineups with all null plots. Overall, we collected `r vi_survey  %>% filter(!attention_check) %>% nrow()` evaluations on `r vi_survey  %>% filter(!attention_check) %>% pull(unique_lineup_id) %>% unique() %>% length()` unique lineups performed by `r vi_survey  %>% filter(!attention_check) %>% count(exp, set) %>% nrow()` subjects throughout three data collection periods.

## Simulating departures from good residuals

### Non-linearity

Data collection period I is designed to study the ability of human subjects to detect the effect of a non-linear term $\boldsymbol{z}$ constructed using Hermite polynomials on random vector $\boldsymbol{x}$ formulated as

\begin{align} \label{eq:nonlinearity-model}
\boldsymbol{y} &= 1 + \boldsymbol{x} + \boldsymbol{z} + \boldsymbol{\varepsilon},\\
\boldsymbol{x} &= g(\boldsymbol{x}_{raw}, 1), \\
\boldsymbol{z} &= g(\boldsymbol{z}_{raw}, 1), \\
\boldsymbol{z}_{raw} &= He_j(g(\boldsymbol{x}, 2)),
\end{align}

\noindent where $\boldsymbol{y}$, $\boldsymbol{x}$, $\boldsymbol{\varepsilon}$, $\boldsymbol{x}_{raw}$, $\boldsymbol{z}_{raw}$ are vectors of size $n$, $He_{j}(.)$ is the $j$th-order probabilist's Hermite polynomials, $\varepsilon \sim N(\boldsymbol{0}, \sigma^2\boldsymbol{I}_n)$, and $g(\boldsymbol{x}, k)$ is a scaling function to enforce the support of the random vector to be $[-k, k]^n$ defined as

```{=tex}
\begin{equation} \label{eq:scaling-function}
g(\boldsymbol{x}, k) = (\boldsymbol{x} - min(\boldsymbol{x}))/max(\boldsymbol{x} - min(\boldsymbol{x}))2k - k, \quad \text{for} \quad k > 0. 
\end{equation}
```

According to @abramowitz1964handbook, Hermite polynomials were initially defined by @de1820theorie, but named after Hermite [@hermite1864nouveau] because of the unrecognisable form of Laplace's work. When simulating $\boldsymbol{z}_{raw}$, function `hermite` from the R package `mpoly` [@mpoly] is used to generate Hermite polynomials.

The null regression model used to fit the realizations generated by the above model is formulated as

```{=tex}
\begin{equation} \label{eq:null-model}
\boldsymbol{y} = \beta_0 + \beta_1 \boldsymbol{x} + \boldsymbol{u},
\end{equation}
```
\noindent where $\boldsymbol{u} \sim N(\boldsymbol{0}, \sigma^2\boldsymbol{I}_n)$.

Since $z = O(x^j)$, for $j > 1$, $z$ is a higher order term leaves out by the null regression, which will lead to model misspecification.

Visual patterns of non-linearity are simulated using four different orders of probabilist's Hermite polynomials ($j = 2, 3, 6, 18$).  (A summary of the factors is given in Table \ref{tab:model-factor-table}.) The values of $j$ is chosen so that distinct shapes of non-linearity are included in the residual plot. These include "U", "S", "M" and "triple-U" shape as shown in Figure \ref{fig:different-shape-of-herimite}. A greater value of $j$ will result in a curve with more turning points. It is expected that the "U" shape will be the easiest one to detect because complex shape tends to be concealed by cluster of data points.

Figure \ref{fig:example-poly-lineup} demonstrates one of the lineups used in non-linearity detection. This lineup is produced by the non-linearity model with $j = 6$. The data plot location is $2^3 - 4$. All five subjects correctly identify the data plot from this lineup.

```{r results='asis'}
data.frame(j = c("2", "3", "6", "18", ""),
           sigma = c("0.25", "1.00", "2.00", "4.00", ""),
           a = c("-1", "0", "1", "", ""),
           b = c("0.25", "1.00", "4.00", "16.00", "64.00"),
           n = c("50", "100", "300", "", ""), 
           x_dist = c("Uniform", "Normal", "Skewed", "Discrete uniform", "")) %>%
  knitr::kable(col.names = NULL,
               align = "rr|rr|rc",
               escape = FALSE,
               format = "latex",
               booktabs = TRUE, 
               caption  = 'Levels of the factors used in data collection periods I, II, III.', 
               label = "model-factor-table" ) %>%
  kableExtra::kable_styling(latex_options = "scale_down") %>%
  kableExtra::add_header_above(c("Poly Order (sym1)", #"Order of Hermite polynomial\n(sym1)", 
                                 "SD (sym3)", #"Standard deviation of polynomial model\n(sym3)",
                                 "Shape (sym4)", # "Shape of heteroskedasticity\n(sym4)", 
                                 "Ratio (sym5)", # Variance factor of heteroskedasticity model\n(sym5)",
                                 "Size (sym6)", 
                                 "Distribution of fitted values")) %>% #"Sample size\n(sym6)")) %>%
  kableExtra::add_header_above(c("Non-linearity" = 2, 
                               "Heteroskedasticity" = 2, "Common" = 2)) %>%
  as.character() %>%
  gsub("sym1", "$j$", .) %>%
  gsub("sym3", "$\\\\sigma$", .) %>%
  gsub("sym4", "$a$", .) %>%
  gsub("sym5", "$b$", .) %>%
  gsub("sym6", "$n$", .) %>%
  cat()
```

```{r different-shape-of-herimite, fig.height = 2, fig.cap = "Polynomial forms generated for the residual plots used to assess detecting non-linearity. The four shapes are generated by varying the order of polynomial given by $j$ in $He_j(.)$."}

# Facet label
shape_labels <- c('He[2]:"U"', 'He[3]:"S"', 'He[6]:"M"', 'He[18]:"triple-U"')

# Data for shape 1
dat_shape_1 <- poly_model(shape = 1, 
                          x = {
                            raw_x <- rand_uniform(-1, 1);
                            closed_form(~stand_dist(raw_x))
                            }, 
                          sigma = 0.05)$gen(300) %>%
  mutate(shape = shape_labels[1])

# Generate data for shape 2, 3 and 4. Reuse x and e.
map_df(2:4, function(shape) {
  poly_model(shape = shape, 
                  x = {
                    raw_x <- rand_uniform(-1, 1); 
                    closed_form(~stand_dist(raw_x))
                    }, 
                  sigma = 0.05)$
  gen(300, computed = select(dat_shape_1, x, e)) %>%
  mutate(shape = shape_labels[shape])
}) %>%

  # Combined with data for shape 1
  bind_rows(dat_shape_1) %>%
  mutate(shape = factor(shape, levels = shape_labels)) %>%
  VI_MODEL$plot(remove_axis = TRUE, remove_grid_line = TRUE, theme = theme_light()) +
  facet_wrap(~shape, scales = "free", labeller = label_parsed, ncol = 4) +
  xlab("Fitted values") +
  ylab("Residuals")
```

```{r different-sigma, fig.height = 2, fig.cap = 'Examining the effect of $\\sigma$ on the signal strength in the non-linearity detection, for $n=300$, uniform fitted value distribution and the "U" shape. As $\\sigma$ increases the signal strength decreases, to the point that the "U" is almost unrecognisable when $\\sigma=4$.'}

# Generate data for sigma 0.5
dat_sigma_05 <- poly_model(shape = 1, 
                           x = {
                             raw_x <- rand_uniform(-1, 1);
                             closed_form(~stand_dist(raw_x))
                             }, 
                           sigma = 0.5)$gen(300) %>%
  mutate(sigma = "sigma: 0.5")

# Generate data for other sigma
map(c(1, 2, 4), function(sigma) {
  poly_model(shape = 1, 
                  x = {
                    raw_x <- rand_uniform(-1, 1); 
                    closed_form(~stand_dist(raw_x))
                    }, 
                  sigma = sigma)$
  gen(300, computed = select(dat_sigma_05, x)) %>%
  mutate(sigma = paste0("sigma: ", sigma))
}) %>%

  # Combined with data for sigma 0.5
  bind_rows(dat_sigma_05) %>%
  VI_MODEL$plot(remove_axis = TRUE, remove_grid_line = TRUE, theme = theme_light()) + 
  facet_wrap(~sigma, ncol = 4, scales = "free", labeller = label_parsed) +
  xlab("Fitted values") +
  ylab("Residuals")
```

```{r example-poly-lineup, fig.height = 7, fig.width = 7, fig.cap = "One of the lineups containing non-linearity patterns used in data collection period I. Can you spot the most different plot? The data plot is positioned at $2^3 + 1$."}
VI_MODEL$plot_lineup(vi_lineup$poly_24$data, 
                     remove_grid_line = TRUE, 
                     theme = theme_light(),
                     remove_axis = TRUE)
```

### Heteroskedasticity

Data collection period II is designed to study the ability of human subjects to detect the appearance of a heteroskedasticity pattern under a simple linear regression model setting:

```{=tex}
\begin{align} \label{eq:heter-model}
\boldsymbol{y} &= 1 + \boldsymbol{x} + \boldsymbol{\varepsilon},\\
\boldsymbol{x} &= g(\boldsymbol{x}_{raw}, 1),\\
\boldsymbol{\varepsilon} &\sim N(\boldsymbol{0}, 1 + (2 - |a|)(\boldsymbol{x} - a)^2b \boldsymbol{I}), 
\end{align}
```

\noindent where $\boldsymbol{y}$, $\boldsymbol{x}$, $\boldsymbol{\varepsilon}$ are vectors of size $n$ and $g(.)$ is the scaling function defined in Equation \ref{eq:scaling-function}.

The null regression model used to fit the realizations generated by the above model is formulated exactly the same as Equation \ref{eq:null-model}.

For $b \neq 0$, the variance-covariance matrix of the error term $\boldsymbol{\varepsilon}$ is correlated with the predictor $\boldsymbol{x}$, which will lead to the presence of heteroskedasticity. Visual patterns of heteroskedasticity are simulated using three different shapes ($a$ = -1, 0, 1). (A summary of the factors can be found in Table \ref{tab:model-factor-table}.)

>>>>

---
title: |
  Appendix: A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol
author:
  - name: Weihao Li
    affil: a
    email: weihao.li@monash.edu
  - name: Dianne Cook
    affil: a
    email: dicook@monash.edu
  - name: Emi Tanaka
    affil: a
    email: emi.tanaka@monash.edu
  - name: Susan VanderPlas
    affil: b
    email: susan.vanderplas@unl.edu
affiliation:
  - num: a
    address: |
      Department of Econometrics and Business Statistics, Monash University, Clayton, VIC, Australia
  - num: b
    address: |
      Department of Statistics, University of Nebraska, Lincoln, Nebraska, USA
bibliography: paper.bib
header-includes: |
  \usepackage{lscape}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
output: rticles::tf_article
editor_options: 
  chunk_output_type: console
---

@@
